Classless Java

### 2016-03-22 OOPSLA16/sections/Appendix_Formalization.tex
Some related work reorganization

>>>>

\section{Expanded related work}\label{sec:ERW}

\paragraph{Meta-Programming Competes with Language Extensions}
The most obvious solution to adding features to a language is via syntactic
extensions. Syntactic extensions are often implemented as
desugarings to the base language. For example,
the Scala compiler was extended to directly support XML syntax.
However, when syntactic extensions are independently created it is hard
to combine multiple extensions into one.
SugarJ~\cite{erdweg11sugarj} is a Java-based extensible language that
aims at making the creation and composition of syntactic sugar
extensions easy, by allowing programmers to extend Java with custom
features (typically for DSLs). However SugarJ's goals are different
from language tuning: SugarJ aims at creating and composing new syntax;
whereas language tuning merely reinterprets existing syntax. It is
clear that reinterpreting existing syntax only can be limiting
for some applications. However, when language tuning is possible
it has the advantage that existing tools for the language work
out-of-the-box (since the syntax is still the same); and composition
of independently developed tunings is straightforward.

Scala-Virtualized~\cite{Rompf2012scalavirtualized} is an extension to Scala,
which allows blending shallow and deep embedding of DSLs. It redefines
some of Scala's language constructs to method calls,
which can be overridden by DSL implementer. Thus Scala-Virtualized
can also reinterpret syntax, and be seen as a form of language tuning.
However, although many Scala's language constructs are supported,
not all language constructs can be virtualized.

%These extensions are admittedly useful, however they are essentially
%creating a fork of the language, and rarely the new fork gains enough
%traction to become the main language release. % On this
% topic we mention

When the base language has a flexible enough syntax and a fast and
powerful enough reflection mechanism, we may just need to play with
operator overloading and other language tricks to discover that the
language feature we need can be expressed as a simple library in our
language. An example of this is SQLAlchemy~\cite{bayer2012sqlalchemy} in
Python, which uses operator overloading to dynamically turn normal python expressions into database queries
without requiring any syntactic extensions to Python.
Java-like languages tend to sit in the middle of two extremes:
libraries can not influence the type system, so many solutions valid
in Python or other dynamic languages are not applicable, or have the
cost of losing type-safety.

In Java-like languages compile time code generation comes at the rescue: if, for a certain feature
(\mixin in our case), it is possible to use the original language syntax to
\emph{express/describe} any specific instantiation of such feature (annotating a
class and providing getters), then we can insert in the compilation process a tool
that examines and enriches the code before compilation. No need to modify the
original source (for example we can work on temporary files).
Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process via annotation
processing. In this way
there is no need of temporary files, and there is a well-defined way of putting
multiple extensions together.

Other languages offer even stronger support for safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time. They generate code that is
transparently integrated in the program that is being
generated/processed/compiled. In particular, MetaFjig offers a property called
\emph{meta-level-soundness}, ensuring by construction that library code (even if wrong or unreasonable) never generates ill-typed code. This is roughly equivalent to
\textit{Self coherence}, that we have to manually prove.
Since MetaFjig is not working on annotated classes, there is not a ``semantic with/without annotations''. Our \mixin tuning theorem does not make sense in such context.

\paragraph{Formalization of Java 8}
We provide a simple formalization for a subset of Java including
default/static interface methods and object initialization literals
(often called anonymous local inner classes).  A similar formalization
was drafted by Goetz and Field~\cite{goetz12fdefenders} to formalize
defender (default) methods in Java. In their formalization, classes
and interfaces can have only one method \Q@m()@ without arguments, so
as to simplify method overloading and renaming. Classless Java is more
general, as it supports multiple methods with arguments, it supports
static methods, and features such as multiple inheritance of
interfaces and reabstraction of default methods are also modeled.

\section{Interaction of Interface Methods with Interface Composition}
\bruno{I don't think this section makes alot of sense in the context 
of the new paper! I think perhaps some of this could be mentioned 
in related work.

In my opinion, instead of this section, we could have one section 
that demonstrates our implementation and its advantages.

\begin{itemize}

\item Show a diagram of how the approach works?

\item Syntax and type system reuse; (good for error messages, for example)
\item On-the-fly AST rewriting (means no new files are created, and functionality
is immediatelly available for IDE's for example); this is better than conventional 
annotation processors.
\item Eclipse integration; code completion.
\item No integration issues in typical pre-processors; (the process is transparent; 
no need for an external tool that generates code in a first step).
\item Separate compilation (which relates to the properties?)

\end{itemize}

}
Before formalizing Classless Java and object interfaces, it is helpful
to informally discuss the behaviour of Java8
interfaces, in contrast with conventional trait models.

%We show several interesting cases
%and summarize compilation result of method composition into 3 categories:
%conflict error, accepted (both abstract) and conservative error. Meanwhile we
%also show different composition handling mechanism among traits, javac and ECJ.

\subsection{Methods in Java 8 Interfaces}
In Java 8 interfaces there are three types of methods: abstract, default, and static methods. Default and static methods were not allowed in interfaces in previous versions of Java.

\paragraph{Static methods} are handled in a very clean way: they are visible only in
  the interface in which they are explicitly defined. This means the following code
  is ill-typed.
\begin{lstlisting}
interface A0 { static int m(){return 1;} }
interface B0 extends A0 {}
... B0.m()//ill typed
\end{lstlisting}
This is different from the way static methods are handled in classes. Here
static methods have simply no interaction with interface
composition (\Q@extends@ or \Q@implements@).

\paragraph{Abstract method} composition is accepted when there exists a most specific one.
  For example, here method \texttt{Integer m()} from \texttt{B1} is visible in \Q@C1@.
\begin{lstlisting}
interface A1 { Object m(); }
interface B1 { Integer m();}
interface C1 extends A1,B1 {} //accepted
\end{lstlisting}

\paragraph{Default methods} conflict with any other default or abstract method. For
  example the following code is rejected due to method conflicts.
\begin{lstlisting}
interface A2 { default int m() {return 1;}}
interface B2 { int m(); }
interface C2 { default int m() {return 2;}}
interface D2 extends A2,B2 {} //rejected due to conflicting methods
interface E2 extends A2,C2 {} //rejected due to conflicting methods
\end{lstlisting}
Note how this is different from what happens in most trait models, where \Q@D2@
would be accepted, and the implementation in \Q@A2@ would be part of the
behavior of \Q@D2@.

\paragraph{Resolving conflicts:}
A method in the current interface wins over any method in its
super-interfaces, provided that the method
is the most specific one. This method also overrides conflict due to
inheritance. For example, the following code is accepted, but would be rejected
(see before) if the method \Q@m@ was not redefined in \texttt{D3} and
\texttt{E3}.
\begin{lstlisting}
interface D3 extends A2,B2 { int m(); } //accepted
interface E3 extends A2,C2 { default int m(){return 42;} } //accepted
\end{lstlisting}

\subsection{Classifying Outcomes of Interface Composition}
%When interfaces are composed and methods with the same name (and
%signature) exist, there are 3 possible outcomes.
%
We now try to classify possible outcomes for composition of methods with the same name (and
signature).
We will use the following (correct) declarations:
\begin{lstlisting}
interface A1 { T m(); }
interface A2 extends A1 { default T m(){ ... } }
interface A3 extends A2 { T m(); }

interface B1 { default T m(){ ... } }
interface B2 extends B1 { T m(); }
interface B3 extends B2 { default T m(){ ... } }
\end{lstlisting}

\noindent
What happens if a new interface \Q@M@ extends one \Q@A@${}_i$ and one
\Q@B@${}_j$?
% Nine representative cases are shown next:\\*
\begin{figure*}[htbp]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & conservative error                     & conflict error      & conservative error       \\ \hline
\textbf{B2}        & both abstract (accepted)     & conservative error       & both abstract (accepted)       \\ \hline
\textbf{B3}        & \textbf{conservative error} &conflict  error       & conservative error      \\ \hline
\end{tabular}
\end{figure*}
%We try to classify the results in the table:
\begin{itemize}
\item \textbf{conflict error} happens when the methods from both interfaces are implemented, which is also an error in most trait models.
\item \textbf{both abstract (accepted)} happens when the methods from both interfaces are abstract, which is also considered correct in all
  trait models.
\item \textbf{conservative error} happens when only one method is implemented
  (leaving another one abstract), which is different from what we would expect in
  a trait model, but is coherent with the conservative idea that a method
  defined in an interface should not silently satisfy a method in another one.
\end{itemize}

\paragraph{A bug:} During our experimentation, we found a bug in ECJ (Eclipse compiler for Java):
the case \textbf{M} extending \textbf{B3} and \textbf{A1} is accepted by
ECJ4.5.1 and rejected by javac.  By email communication with Brian Goetz
(leading Java 8 designer) we have confirmed that the expected behavior is
rejection, hence this is a bug in ECJ. This bug was also reported by
  others and is fixed in the ECJ developer branch, but not released as a stable
  version yet.

### 2016-03-22 OOPSLA16/sections/RelatedWork.tex
Some related work reorganization

\emph{Traits vs Object Interfaces.}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation-driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with state is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
state, while still being quite expressive.  Moreover the ability to support
constructors adds expressivity, which is not available in approaches
such as Scala's traits/mixins.

@@

@@

\paragraph{Meta-Programming Competes with Language Extensions}
The most obvious solution to adding features to a language is via syntactic
extensions. Syntactic extensions are often implemented as
desugarings to the base language. For example,
the Scala compiler was extended to directly support XML syntax.
However, when syntactic extensions are independently created it is hard
to combine multiple extensions into one.
SugarJ~\cite{erdweg11sugarj} is a Java-based extensible language that
aims at making the creation and composition of syntactic sugar
extensions easy, by allowing programmers to extend Java with custom
features (typically for DSLs). However SugarJ's goals are different
from language tuning: SugarJ aims at creating and composing new syntax;
whereas language tuning merely reinterprets existing syntax. It is
clear that reinterpreting existing syntax only can be limiting
for some applications. However, when language tuning is possible
it has the advantage that existing tools for the language work
out-of-the-box (since the syntax is still the same); and composition
of independently developed tunings is straightforward.

Scala-Virtualized~\cite{Rompf2012scalavirtualized} is an extension to Scala,
which allows blending shallow and deep embedding of DSLs. It redefines
some of Scala's language constructs to method calls,
which can be overridden by DSL implementer. Thus Scala-Virtualized
can also reinterpret syntax, and be seen as a form of language tuning.
However, although many Scala's language constructs are supported,
not all language constructs can be virtualized.

%These extensions are admittedly useful, however they are essentially
%creating a fork of the language, and rarely the new fork gains enough
%traction to become the main language release. % On this
% topic we mention

When the base language has a flexible enough syntax and a fast and
powerful enough reflection mechanism, we may just need to play with
operator overloading and other language tricks to discover that the
language feature we need can be expressed as a simple library in our
language. An example of this is SQLAlchemy~\cite{bayer2012sqlalchemy} in
Python, which uses operator overloading to dynamically turn normal python expressions into database queries
without requiring any syntactic extensions to Python.
Java-like languages tend to sit in the middle of two extremes:
libraries can not influence the type system, so many solutions valid
in Python or other dynamic languages are not applicable, or have the
cost of losing type-safety.

In Java-like languages compile time code generation comes at the rescue: if, for a certain feature
(\mixin in our case), it is possible to use the original language syntax to
\emph{express/describe} any specific instantiation of such feature (annotating a
class and providing getters), then we can insert in the compilation process a tool
that examines and enriches the code before compilation. No need to modify the
original source (for example we can work on temporary files).
Java is a particular
good candidate for this kind of manipulation since it already provide ways to
define and integrate such tools in its own compilation process via annotation
processing. In this way
there is no need of temporary files, and there is a well-defined way of putting
multiple extensions together.

Other languages offer even stronger support for safe code manipulation:
Template Haskell~\cite{sheard2002template}, F\# (type providers)~\footnote{http://research.microsoft.com/fsharp/} and MetaFjig (Active Libraries)~\cite{servetto2010metafjig}
all allow to execute code at compile time. They generate code that is
transparently integrated in the program that is being
generated/processed/compiled. In particular, MetaFjig offers a property called
\emph{meta-level-soundness}, ensuring by construction that library code (even if wrong or unreasonable) never generates ill-typed code. This is roughly equivalent to
\textit{Self coherence}, that we have to manually prove.
Since MetaFjig is not working on annotated classes, there is not a ``semantic with/without annotations''. Our \mixin tuning theorem does not make sense in such context.

\paragraph{Formalization of Java 8}
We provide a simple formalization for a subset of Java including
default/static interface methods and object initialization literals
(often called anonymous local inner classes).  A similar formalization
was drafted by Goetz and Field~\cite{goetz12fdefenders} to formalize
defender (default) methods in Java. In their formalization, classes
and interfaces can have only one method \Q@m()@ without arguments, so
as to simplify method overloading and renaming. Classless Java is more
general, as it supports multiple methods with arguments, it supports
static methods, and features such as multiple inheritance of
interfaces and reabstraction of default methods are also modeled.

>>>>

\paragraph{Traits vs Object Interfaces.}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation-driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with state is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
state, while still being quite expressive.  Moreover the ability to support
constructors adds expressivity, which is not available in approaches
such as Scala's traits/mixins.\footnote{Scala traits have
  no constructor, but \Q@var@s and (\Q@lazy@) \Q@val@s can be
  ``pre-initialized''.}

@@

% \footnote{The original
%  notion of traits by Scharli et al. prescribes, among other things,
%  that: 1) a trait provides a set of methods that implement behavior;
%  and 2) a trait does not specify any state variables, so the methods
%  provided by traits do not access state variables directly. Java 8
%  interfaces follow similar principles too. Indeed, a detailed
%  description of how to emulate trait-oriented programming in Java 8
%  can be found in the work by Bono et al.~\cite{bono14}. The Java 8
%  team designing default methods, was also fully aware of that
%  secondary use of interfaces, but it was not their objective to model
%  traits: ``The key goal of adding default methods to Java was
%  "interface evolution", not "poor man's
%  traits"''~\cite{goetz13default}. As a result, they were happy to
%  support the secondary use of interfaces with default methods as long
%  as it did not make the implementation and language more complex.}
\marcoT{We could use Scala Traits to encode IB, but the simpler Java8
  model is a better starting point: scala traits can declare \Q@def, val, var@ and
  \Q@lazy val@ However, while \Q@def@s can be overridden by
  \Q@val@/\Q@var@s, \Q@val@/\Q@var@s can not be overridden.
   A\Q@val@/\Q@var@ member is predestined to be
  instance state, and this is in sharp contrast with our model where
  the state is only a private detail of instances, and do not
  contaminate interface/trait/classes.
  }

@@
